Scaling up a GF grammar for free parsing

### 2011-12-13 doc/report/report.tex
breport: better results-sections, mapping, saldo

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inhereted ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has succesfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in wich GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{molto} for online translation between up to 15 language.\\

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large-scale treebank,
Talbanken \cite{talbanken} was perfect for our purpose.
%It was created in the 1970s at Lund University.
In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser \cite{malt}
and was then redistributed in a more modern version, Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
XML-formats
where the trees had been made deeper and more detailed. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

@@

\section{Related work}
\label{sec:related}
There has been much research about parsing Swedish. One of the most well known
parser is the data-driven Malt parser \cite{malt}, which has been trained on 
Talbanken. 
There are also a number of grammar-based parsers, although none is still freely avaible.
CassSwe \cite{casswe} is a cascaded finite state parser giving a syntactic
analysis.
The Swedish Constraint Grammar 
(Birn,1998) gives a shallow syntactic analysis.
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
(Tapanainen and Järvinen,1997), an extension of the Constraint Gramman
formalism, and produces a dependecy structure focusing on finding the nominal
arguments. \\
%for the  finding subject, objects, suubject complements (SC) and
%to link them to their proper regents (main verbs) : over 90\% both precision and recall.

@@

The LinGO Grammar Matrix \cite{matrix}, is a starter-kit for building Head-Driven Phrase
Structure Grammars \cite{hpsg} (HPSG) providing comptability with tools for
parsing, evaluation, semantic representations etc and  
translation is supported by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. \\
There is a collection of grammars implemented in this framework, giving broad-coverage
description to ie.  %broad-coverage
English, Japanese and German. %, Greece, Spanish etc. 
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandiavian, while Norsource (Hellan, 2003) describes Norweigan. A Swedish version
was based upon this (SweCore, Ahrenberg) which covers the morphology and some
differences between Swedish and Norweigan. Further, there is the BiTSE \cite{stymne}
grammar, also implemented using the Lingo Matrix,
which focuses on verb frames and translation of those to English. \\

@@

The Swedish verison \cite{gamback} of the Core Language Enigne (CLE) \cite{cle}
gives a full syntacitc analysis as well as semantics represented in xx. A
translation for conversations?? to English was implemented. The work was further developed in the spoken
language blabla (xx,xx), but is no longer available. The coverage of the Swedish
CLE is also reported to be very limited (ref).\\

@@

\section{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}) and numerals(\emph{nl}) etc.
~~are assumed to be known already~~,
and are usually not kept in the standard lexicon, but in modules for the basic
structures of the language. Categories involving several words(\emph{vbm,nnm})
need to be handled differently, usually as idioms. Pronouns(\emph{pn}) are not  
analysed the same way, see section \ref{sec:swegf}. Some experiments on finding
the right GF-category for those have been done, see ~~~~~\ref{sec:gf.quant}.
Six categories were imported: \\
>> english examples
\begin{tabular}{l|lll}
& Saldo tag & GF category & Example \\
\hline
 Adverb & \textbf{ab} &\textbf{Adv} & ofta \\
 Adjective&\textbf{av} &    \textbf{A} & gul\\
 Noun & \textbf{nn} &\textbf{N} & hus\\
 Verb & \textbf{vb} &\textbf{V} & springa\\
 Reflexive verbs  &\textbf{vbm}& \textbf{V} & raka sig\\
 Verbs with particles &\textbf{vbm}& \textbf{V}  &  piggna till\\
\end{tabular}\\

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
One reason for why words of the categories listed above could not be imported
is that Saldo does not contain information about ~~word forms that are not used~~.
In some cases those forms are needed by the smart paradigm. 
This was a common reason why nouns could not be imported. 
Consider for example
the plural tantum % \cite[44]{SAG}
noun ``glasögon" (``glasses"). Since the smart paradigm
requires a singular form, this lemma could not be added to the lexicon. ``Glasögon" could
be added manually, by using the ~~somewhat incorrect,annan betydelse~~ singular form ``glasöga" (``glass-eye").
The same problem occurred for verbs that always are used in passive form:
>> not passive!! deponent, s-verbs!!
``andas" (``breathe"), ``frodas" (``flourish"), ``lyckas" (``succeeed"). This
was the case for close to 80\% of the failing verbs of type \verb_vb_.

In some cases the smart paradigms could not generate the correct declination,
although this happened seldom. 
>> exapmles: % frysa,fisa passive preterium

The resulting dictionary contains more than 100 000 entries, compared to the total size
of Saldo this is approximately 80\%. When testing the coverage of Talbanken,
we found that around 3000 words were still missing (excluding the one tagged as
numerals and proper names, but some names are still there). In addition, there were 1817 words
>> give reasons! should look good, not bad!
which were given different labels in GF than in Talbanken. Many of those are
acceptable, such as the ones in table \ref{tab:saldodiff} and reflects the difference 
made in the analyses, but others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{lll}
Word & Talbanken tag & GF category \\
\hline
måste & MVPS          & VV \\
allting & POTP & N \\
få & POZP & Det \\
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}
Among the words that were completely missing, more than 2000 was found to be nouns,
many of which were compounds, and this is hard because ...~~~

The lexicon does not contain any valency information since Saldo does not give any such
information. Valencies are however crucial for GF, and by using the results from the mapping
(see section \ref{sec:Mapping}) it should be possible to extract this information about this.
>> about what? how? strategy?
%The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
%have any information. The earlier explained techniques may be used for this in Talbanken.
%To get even bigger material, use Korp, but then no guarantee.

@@

\chapter{Mapping}
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.
>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
>> mamba very used and tested for swedish, standard.
>> focus on what we DO cover.
\label{sec:Mapping}
Talbanken contains much valuable information about phrase structures and word 
usage. It is analysed with the MAMBA annotation scheme (Telemann, 1974) which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
commonly used, and hence very well-tested, for describing Swedish.
The deepened analyse in Talbanken05 includes the lexical MAMBA layer. 
One part of this project has focused translating trees from Talbanken to GF.
We hence get a comparision between the two annotaitons and at the same time
we get ways to extract and translate information to GF notation. We have
constructed a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence \emph{``The cat on the car gets bigger"} and the result of translating it to GF.
During the translation all the information given in the abstract tree
(\ref{pic:gftree}) needs to be extracted. We need to know that \emph{``katten"} is a noun
used in definite form (\verb-DetArt-), singular (\verb-NumSg-). Even though the
notation is changed by the transformation, the original analyse from Talbanken,
that ``på bilen" forms a subtree which is attatched to ``katten" for example,
is preserved.
%produces a \verb-NP- from ``katten" this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten
Both the POS tags as well as the syntactic information are needed.

\begin{figure}[h]
\hspace{-10mm}
\subfloat[Talbanken tree]{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gftree}\includegraphics[width=60mm]{gfTree.pdf}}
\subfloat[GF parse tree]{\label{pic:gfptree}\includegraphics[width=50mm]{gfPTree.pdf}}
\caption{Trees for the sentence ``Katten på bilen blir större".}
\label{fig:translationtrees}
\end{figure}

The mapping connects GF with another annotations and gives us means to evaluate
our parser. By both parsing a Talbanken sentence and transforming its annotated tree, we can
easily inspect if the results are equal.
The mapping also makes it clear which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities, see section
\ref{sec:future}, of how often a function is used.\\
%missing words. Lexical aquisation, probabilities, see section .
%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools. % Those can be enhanced if they are given more data. 
%The translated trees enable us to extract those probabilities. %The mapping gives us a source for this.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.

A translation of the English Penn Treebank to the GF syntax has previously been done \cite{gfpenn}.
By starting this from program, we now have
a translation that works for the differently annotated Talbanken. The differences between
Swedish and English syntax also meant that many rules had to be changed.\\

The mapping works for Talbanken05 in Tiger XML-format.
There are three kinds of tags: categories, edge labels and POS tags. 
The categories give syntactic information such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give similar information with tags as \verb|SS| for subject, 
\verb|OO| for object. There are more than 300 different pos tags. The high
number is due to the detailedness, for example there are almost 50 tags for
nouns, excluding proper names. The tags show definiteness, case, whether the
word is a compound. Some words also have their own tags.\\

@@

The translation works section-wise, ie. if it finds a tree of category \verb|S|,
it will start to look for a subject \verb|SS| and a finite verb \verb|FV|. If it
can find those, they can be combine with the GF function \verb|PredVP|.
At the lowest level in the tree we look up the word in the GF lexicon.
It figure \ref{pic:tbtree}, the verb is followed by an adjective as a present
participle, \verb|SP|. GF admits \emph{``bli"} to have type \verb|VA|, a verb
with an adjective as complement, and the translator can apply the GF function
\verb|ComplVA| to combine the verb with its object. \\
It is not always the case that we first find the subject, then the verb and 
finally the objects. Often the order is Verb-Subject-Object or Object-Subject-Verb. \\
During the mapping process, the translator also needs to find out whether the sentence is
negated and what tense is used. If it finds two trees of type, say, \verb|VP| but does not
have any rules for combining them, the \textit{meta function} \verb-?- will be used. 
Figure xx shows an example of this.\\

@@

%Description of most important tags from Talbanken, their translation? appendix?
\subsubsection{Differences in the notation}
Some structural differences between GF and Talbanken were made obvious during the
translation. \\
%Since the translation is done word by word, ambiguities may confuse the translator.\\
%shown during the translation.
%cause some translation problems.
In Talbanken the valency information is given implicitly by the complements
of a word. If a verb is followed by an object, \verb-OO-, which contains a
\verb-S-, we can conclude that this is a sentence-complemnt verb.
In GF, the valency information is given for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and from this we know 
that the function \verb-ComplVS- should be used to combine the verb and with a 
sentence.
A similar issue arises form the use of verb particles. A verb with a
particle is considered as one constituent in GF, which means that the particle
has no effect on the abstract tree. Its presence is announced by the verb
itself, while Talbanken treats them as separated units.
%In Talbanken, verbs and particles are treated as separate units,
%and the particle is marked with the tag \verb-PL- and attached to. 
%Talbanken has no valence information while this is crucial for GF.
%The grammar requires a adjective-complement verb,
%\verb|VA|, to be used with one adjectival phrase. For \emph{``bli"} in example
%\ref{pic:tbtree}, we know that we need to find an adjective phrase.
%%know what kind of verb we are looking for
%%when consulting the GF lexicon.
%If the GF lexicon assings different types to verb, the correct type 
%should be showed by the complements.
%This way %\enumsentence{Han gillar det\\He likes it} 
The mapping cannot depend on the valencies given in GF only, however, 
since the distinction between different types of verbs is not always clear in Swedish.
The grammar rules becomes too strict for accepting most Talbanken sentences.
For example, most transitive verbs can also be used as intransitive. Both sentences in
example \ref{ex:mappTransitive} are grammatically correct.
\enumsentence{
\shortexm{11}
{Han& sitter& och &läser.}
{He& sits& and &reads.}{}
{Han& sitter& och& läser& en& bok.}
{He& sits& and& reads& a& book.}{}}
\label{ex:mappTransitive}
See section \ref{sec:futureValency} for a longer discussion about this.\\
%the Talbankenwill allow words to be used in other ways than their
%GF valencies prescribes.
Another problem can be illustrated by sentence \ref{ex:mappMan}, containing
the generic pronoun \emph{``man"}.
\enumsentence{
\shortex{3}{Man & uppskattar & dem}
{One & appreciates & them}
{``They are appreciated"}}\label{ex:mappMan} %  (s216).
In Talbanken, \emph{``man"} is simply a personal pronoun, 
in GF it is represented by using the function \verb|GenericCl|, which 
turns a \verb-VP- into a clause and has no
subject visible in the abstract tree. See figure \ref{fig:mappMan}.
\begin{figure}[h]
\centering
\includegraphics[width=80mm]{man.pdf}
\caption{Abstract tree for ``Man uppskattar dem".}
\label{fig:mappMan}
\end{figure}
When translating a senctence like this, it is thus not possible to simply glue the
subject to the verb with \verb|PredVP|.\\
%Therefore, the translator must change the type of the 
%sentence when inside the SS.

>>>>

%>>the scope is controlled language, but possible ambiguities (discuss this in future
%>>work, saldo, conclusion). Ambiguities: from multilinguality, or inherited:
%>>syntactical (man with the telescope), lexical (youSg/Pl, växa).
%>> have ambiguity section. 
\newpage %as long as english picture bråks.
The scope of GF grammars is usually controlled language. By
restricting the coverage, the number of ambiguities
can be limited and we can ensure that the semantics is preserved during
translation. Inhereted ambiguities and the ones arising from multilinguality will
of course remain, but can be controlled more easily. %Example of domainspecific word?
The use of controlled language thus gives a possibility of good and reliable translation
and GF has succesfully been used in project like %Some examples of projects using GF are 
WebAlt \cite{webalt}, a project which aims to develop language independent
material for mathematics problems, and the formal software verification tool
KeY \cite{key}, in wich GF is used for translating specification to natural language.
%>> key, talk, hats, bom (see references in Ramonas thesis).
%>> use talk here and move key to other place??
The framework is also used in the
European project MOLTO \cite{Molto} for online translation between up to 15 language.\\

@@

\section{Talbanken}
\label{sec:talbanken}
For testing and evaluation of the grammar and lexicon, we needed to be able to
compare them against a reliable source.
Being a freely available, manually annotated, large-scale treebank,
Talbanken \cite{talbanken} was perfect for our purpose.
%It was created in the 1970s at Lund University.
In total it consists of four
parts. Two of them are transcribitions of spoken language, one a collection of
text written by high school students, and one, section P,
consists of professionally written Swedish gathered from newspapers, brochures and textbooks.\\
Talbanken was also used to train the Swedish version of the Maltparser \cite{malt}
and was then redistributed in a more modern version, Talbanken05 \cite{talbanken05}.
%This work was done by Nivre, Hall and Nilsson in 2005.
It is released in Malt\footnote{http://w3.msi.vxu.se/~nivre/research/MaltXML.html} 
and Tiger\footnote{http://www.ims.uni-stuttgart.de/projekte/TIGER/TIGERSearch/doc/html/TigerXML.html},
XML-formats
where the trees had been made deeper and more detailed while still containing
the lexical MAMBA layer. \\
In this project, as well as for the Malt parser, the more than 6000 sentences
in section P has been used.
The treebank has served as an inspiration and an evaluation source throughout this
project. An automatic mapping between its trees and the abstract trees from GF has been
done, which will be explained in section \ref{sec:Mapping}.
%Differences in analyse, do we want a similar? 
%Uses of Talbanken in project, mapping to evaluate and test.  
%Look at output from Maltparser, could also comparing our parse trees to this.

@@

\section{Related work}
\label{sec:related}
%There has been much research about parsing Swedish. INTRO!!
A well known parser is the data-driven Malt parser \cite{malt}, which has been trained on 
Talbanken. 
There are also a number of grammar-based parsers, although none is still freely avaible.
The cascaded finite state parser CassSwe \cite{casswe} and
The Swedish Constraint Grammar (Birn,1998) 
give syntactic analyses. %birn: shallow
Swedish FDG (Voultanien,2001) uses the Functional Dependency Grammar
(Tapanainen and Järvinen,1997), an extension of the Constraint Gramman
formalism, and produces a dependecy structure focusing on finding the nominal
arguments. \\
%for the  finding subject, objects, suubject complements (SC) and
%to link them to their proper regents (main verbs) : over 90\% both precision and recall.

@@

The LinGO Grammar Matrix \cite{matrix}, is a starter-kit for building Head-Driven Phrase
Structure Grammars \cite{hpsg} (HPSG) providing comptability with tools for
parsing, evaluation, semantic representations etc.
Translation is supported by using Minimal Recursion
Semantics \cite{mrs} as an interlingua. \\
There is a collection of grammars implemented in this framework, giving broad-coverage
description to %broad-coverage
English, Japanese and German. %, Greece, Spanish etc. 
The Scandinavian Grammar Matrix \cite{scandmatrix} covers common parts of
Scandiavian, while Norsource (Hellan, 2003) describes Norweigan. A Swedish version
was based upon this (SweCore, Ahrenberg) covering the morphology and some
differences between Swedish and Norweigan. Further, there is the BiTSE \cite{stymne}
grammar, also implemented using the Lingo Matrix,
which focuses on describing and translating verb frames.\\ % and translating them to English. \\

@@

The Swedish verison of the Core Language Enigne (CLE) \cite{gamback} % \cite{cle}
gives a full syntacitc analysis as well as semantics represented in xx. A
translation for conversations?? to English was implemented. The work was further developed in the spoken
language blabla (xx,xx). Unfortunally, it is no longer available. The coverage of the Swedish
CLE is also reported to be very limited (ref).\\

@@

\section{Results}
All words from Saldo were not be imported. There are a number of reasons
for this, the most obvious one is that we do not want all categories from
Saldo in the GF lexicon. Prepositions(\emph{pp}), numerals(\emph{nl}),
personal pronouns etc.
are assumed to be present in the resource grammars and should not be added again.
Saldo contains may other pronouns,
but they are not analysed the same way in GF (see section \ref{sec:swegf}) and
can thus not simply be extracted -- we first need to find their correct GF category.
Some experiments on finding
the right GF-category for those have been done using Talbanken, see p. \pageref{sec:gf.quant}.

%and are usually not kept in the standard lexicon, but in modules for the basic
%structures of the language.
Categories involving several words (\emph{vbm, nnm})
are usuallly handled as idioms and should be given in a separate lexicon. In
total six categories were imported: \\

\begin{tabular}{|l|lll|}
\hline
& Saldo tag & GF category & Example \\
\hline
 Adverb & \textbf{ab} &\textbf{Adv} & ofta (\emph{often})\\
 Adjective&\textbf{av} &    \textbf{A} & gul (\emph{yellow})\\
 Noun & \textbf{nn} &\textbf{N} & hus (\emph{house})\\
 Verb & \textbf{vb} &\textbf{V} & springa (\emph{run})\\
 Reflexive verbs  &\textbf{vbm}& \textbf{V} & raka sig (\emph{shave})\\
 Verbs with particles &\textbf{vbm}& \textbf{V}  &  piggna till (\emph{perk})\\
\hline
\end{tabular}\\

%
%There are a number of differences between and saldo Till skillnad från GF,
%saldo does not contain forms that aren't used gf generates all forms. A form
%needed to generate the gf table may not be given in saldo such as singular
%forms for 'byxor'. Also contains idioms, 'hålla huvudet kallt' and has
%different grammatical analyse, a lot of pronouns. Example. 
Most, but not all words of the categories listed above have been imported.
One reason why the importing phase would fail 
is that Saldo, unlike GF, only contains the acctually used word forms.
% does not contain information about ~~word forms that are not used~~.
For technical reasons, the smart paradigm might need forms never used.
%for this is Deponent verbs
%are for example given with their (often non-existing) active form, al
%In some cases those forms are needed by the smart paradigm. 
Consider for example
the plural tantum % \cite[44]{SAG} 
noun \emph{``glasögon"} (\emph{``glasses"}).
The smart paradigm requires a singular form, and since the program couldnot
find this in Saldo, there was no way of adding the lemma to the lexicon. 
When the program failed to import a noun, this was often the explanaiton.
%bbMany nouns fai, \emph{``kläder"} (\emph{``clothes"} etc.
Words of this type could be added manually, for \emph{``glasögon"} by using the
ostensibly correct singular form ``glasöga", although this
has another meaning (``glass-eye").
The same problem occurred for the irregular s-verbs,
(\emph{``synas"} \emph{(``show")} or \emph{umgås} \emph{(``socialize")})
which made up 61.5\% of the failing verbs of type \verb_vb_.\\
%``frodas" (``flourish"), ``lyckas" (``succeeed").
In a few cases the smart paradigms could not generate the correct declination.\\

%although this happened seldom. The passive preteritum is an exapmle of a form 
%oftenly because the one form could not be generated. anbd for passive preteritum could not be guessed.
%that seems hard to guess for some verb.
% >> should we add smth for this?
%\begin{tabular}{3}
%&\textbf{Saldo}& \textbf{GF}
%\textbf{Passive preteritum} & fryses & frys \\
%\end{tabular}
%>> exapmles: % frysa,fisa passive preterium

The resulting dictionary contains more than 100 000 entries, approximately 80\% (retest!)
of the total size of Saldo.
%compared to the total size of Saldo this is .
When testing the coverage of Talbanken,
we found that there are around 2500 wordforms were still missing, excluding the one
tagged as names and numbers. This number may seem very high, but %considering that
4/5 of the wordforms are compounds. We should also take into consideration that 
we cannot automatically find out how many actually stem from the same word, or
how many abbreviation that are present (Talbanken also contains a small number of 
misspellings, probably are enumerated among our missing words). The majority
of the words are only used once.\\

\begin{tabular}{|l|l|}
\hline
Missing words & $\sim$ 2500 wordforms\\
Missing words ignoring compounds & $\sim$ 500 wordforms\\
Missing words used more than once & $\sim$ 500 wordforms\\
Missing words used more than once ignoring compounds & $\sim$ 150 wordforms\\
\hline
\end{tabular}\\

%       The table suggests that many 
%       >> 2537 without IDs and only the ones built of Alpha,only 552 without compounds, still contains abbrev.
%       >> 494 without abbr. spelling errors
%       >> 142 more then one time.
%       numerals and proper names, but some names are still there). In addition, there were 1817 words
%       >> 223 were IDS.++146 HS.
%       >> 426 more than one time, with compounds.
%       >> give reasons! should look good, not bad!
A list of words that were given different labels in GF than in Talbanken has been
composed, consisting of about 1600 entries. Many of those are
acceptable, such as the ones in table \ref{tab:saldodiff} and reflects the difference 
made in the analyses, but others are examples of words that are still missing from
the lexicon.
\begin{figure}[h]
\begin{tabular}{|lll|}
\hline
Word & Talbanken tag & GF category \\
\hline
måste & MVPS          & VV \\
allting & POTP & N \\
få & POZP & Det \\
\hline
\end{tabular}
\caption{}
\label{tab:saldodiff}
\end{figure}

The lexicon does not contain any valency information since Saldo does not give any such
information. Valencies are however crucial for GF, and it remains as future work to
use the results from the mapping (see section \ref{sec:Mapping})
to extract information about how words are normally used and how to automatically build
this into the lexicon. 
%The saddest thing: no valency info. Crucial for GF. Only reflexive and verbs with particles
%have any information. The earlier explained techniques may be used for this in Talbanken.
%To get even bigger material, use Korp, but then no guarantee.

@@

\chapter{Mapping}
\label{sec:Mapping}
>> no subjects or object, analyse is more from the lexicon,
>> compare to others, ask elisabet.
>> diffs: valencies, pp (gf ingen skillnad direkt/indirekt i träd, pga multiling)
>> mamba very used and tested for swedish, standard.
>> focus on what we DO cover.\\
Talbanken contains much valuable information about phrase structures and word 
usage. It is analysed with the MAMBA annotation scheme (Telemann, 1974) which is 
%>> cite mamba? annotation: well intergrated in sw. tradition, descriptive, used
%>> for large scale projects.
%annotation scheme
commonly used, and hence very well-tested, for describing Swedish.
One part of this project has focused translating trees from Talbanken to GF
by constructing a mapping, which 
automatically transforms trees in Talbanken format to GF abstract syntax trees.
We hence get a comparision between the two annotaitons and at the same time
we get methods to extract and translate information to GF notation. 
Figure \ref{fig:translationtrees} shows an example of a visualized Talbanken05 tree
of the sentence \emph{``The cat on the car gets bigger"} and the result of translating it to GF.
Both the POS tags as well as the syntactic information are needed
during the translation, so that all the information given in the abstract tree
(\ref{pic:gftree}) can be extracted. We need to know that \emph{``katten"} is a noun
used in definite form (\verb-DetArt-), singular (\verb-NumSg-). Even though the
notation is translated, the original analyse from Talbanken
is preserved.
\emph{``på bilen"} should still form a subtree attatched to \emph{``katten"}.
%produces a \verb-NP- from ``katten" this means that
%DetCN : Quant                  -> CN    -> NP
%DetCN  (DetQuant DefArt NumSg)    cat_N = katten

\begin{figure}[h]
\hspace{-10mm}
\subfloat[Talbanken tree]{\label{pic:tbtree}\includegraphics[width=60mm]{Talbankentree.pdf}}
\subfloat[GF abstract tree]{\label{pic:gftree}\includegraphics[width=60mm]{gfTree.pdf}}
\subfloat[The corresponding GF parse tree]{\label{pic:gfptree}\includegraphics[width=50mm]{gfPTree.pdf}}
\caption{Trees for the sentence ``Katten på bilen blir större".}
\label{fig:translationtrees}
\end{figure}

The mapping %connects GF with another annotations and 
gives us means to evaluate our parser. By both parsing a Talbanken sentence and
transforming its annotated tree, we can easily inspect if the results are
equal.
Additionally, the mapping clearly shows which grammatical constructions that are still missing
from the GF grammar and shows how the GF analysis differs from the one made
in Talbanken.
If there are words missing from our dictionary, the rich
POS-tags may help us to automatically find the correct declination and add it to the
lexicon. Further, our parser will need probabilities, see section
\ref{sec:future}, of how often a function is used.\\
%missing words. Lexical aquisation, probabilities, see section .
%The mapping gives information about which form a word is
%currently used in and this may be used by the lexical extraction
%tools. % Those can be enhanced if they are given more data. 
%The translated trees enable us to extract those probabilities. %The mapping gives us a source for this.
%Furthermore, the translation makes it easy to identify grammatical constructions
%missing from the GF grammar Another important use of the mapping is evaluation of the parser, which can be
%accomplished by comparing the parse trees and the trees from the transformer.

The mapping is based on 
a translation of the English Penn Treebank to the GF syntax \cite{gfpenn}.
By modifying this program, we now have a translation that works for
the Tiger XML-format of
Talbanken. Adaption was required for the differences in annotation as well as 
syntactic differences between Swedish and English.

%The mapping works for Talbanken05 in Tiger XML-format.
\section{The Talbanken annotation}
Talbanken05 uses three kinds of tags: categories, edge labels and pos-tags. 
While the pos-tags are reserved for words, the categories give syntactic information
for phrases, such as \verb|S|, \verb|NP| and \verb|VP|.
The edge labels give the grammatical functions with tags as \verb|SS| for subject, 
\verb|OO| for object. Nearly 400 different tags are used, of which more than
300 different are pos-tags. The high number is due to the detailedness, for
example there are almost 50 tags for nouns, excluding proper names. The tags
show definiteness, case and whether the word is a compound. Some words also have
their own tags.\\

@@

\section{The translation process}
The translation works section-wise, ie. if it finds a tree of category \verb|S|,
it will start to look for a subject \verb|SS| and a finite verb \verb|FV|. If it
can find those, they can be combine with the GF function \verb|PredVP|.
At the lowest level in the tree we look up the word in the GF lexicon.
It figure \ref{pic:tbtree}, the verb is followed by an adjective as a present
participle, \verb|SP|. GF admits \emph{``bli"} to have type \verb|VA|, a verb
with an adjective as complement, and the translator can apply the GF function
\verb|ComplVA| to combine the verb with its object. \\
It is not always the case that we first find the subject, then the verb and 
finally the objects. Often the order is Verb-Subject-Object or Object-Subject-Verb. \\
During the mapping process, the translator also needs to find out whether the sentence is
negated and what tense is used. If it finds two trees of type, say, \verb|VP| but does not
have any rules for combining them, the \textit{meta function} \verb-?- will be used. 
Figure xx shows an example of this.\\

@@

%Description of most important tags from Talbanken, their translation? appendix?
\subsection{Differences in the notation}
Some structural differences between GF and Talbanken were made obvious during the
translation. \\
%Since the translation is done word by word, ambiguities may confuse the translator.\\
%shown during the translation.
%cause some translation problems.
In Talbanken the valency information is given implicitly by the complements
of a word. If a verb is followed by an object, \verb-OO-, which contains a
\verb-S-, we can conclude that this is a sentence-complemnt verb.
In GF, the valency information is given for each entry in the lexicon.
A sentence-complement verb has the type \verb-VS- and from this we know 
that the function \verb-ComplVS- should be used to combine the verb and with a 
sentence.
A similar issue arises form the use of verb particles. A verb with a
particle is treated as two independent units in Talbanken but in GF they are
considered to be one constituent. % which means that 
The particle has therefore no effect on the abstract tree, its presence is
announced by the verb itself.
%In Talbanken, verbs and particles are treated as separate units,
%and the particle is marked with the tag \verb-PL- and attached to. 
%Talbanken has no valence information while this is crucial for GF.
%The grammar requires a adjective-complement verb,
%\verb|VA|, to be used with one adjectival phrase. For \emph{``bli"} in example
%\ref{pic:tbtree}, we know that we need to find an adjective phrase.
%%know what kind of verb we are looking for
%%when consulting the GF lexicon.
%If the GF lexicon assings different types to verb, the correct type 
%should be showed by the complements.
%This way %\enumsentence{Han gillar det\\He likes it} 
When constructing the abstract trees, we must of course follow and use the
GF analysis. However,
the mapping cannot depend on the valencies given in GF only, 
since the distinction between different types of verbs is not always clear in Swedish;
most transitive verbs can also be used as intransitive. Both sentences in
example \ref{ex:mappTransitive} are grammatically correct, and
the grammar rules are simply too strict for accepting most Talbanken sentences.
\enumsentence{\shortexm{13}
{a. &Han& sitter& och &läser.}
{&\emph{he}& \emph{sits}& \emph{and} &\emph{reads}}{}
{b. &Han& sitter& och& läser& en& bok.}
{&\emph{he}& \emph{sits}& \emph{and}& \emph{reads}& \emph{a}& \emph{book}}{}}
\label{ex:mappTransitive}
See section \ref{sec:futureValency} for a longer discussion about this.\\
%the Talbankenwill allow words to be used in other ways than their
%GF valencies prescribes.

Another annotational difference that turned out to be problematic for the
translation can be illustrated by sentence \ref{ex:mappMan}, containing
the generic pronoun \emph{``man"}.
\enumsentence{
\shortex{3}{Man & uppskattar & dem}
{\emph{one} & \emph{appreciates} & \emph{them}}
{``They are appreciated"}}\label{ex:mappMan} %  (s216).

In Talbanken, \emph{``man"} is simply a personal pronoun, 
in GF it is represented by using the function \verb|GenericCl|, see figure
\ref{fig:mappMan}. \verb|GenericCl| turns a \verb-VP- into a clause and has no
subject visible in the abstract tree. 
\begin{figure}[h]
\centering
\includegraphics[width=80mm]{man.pdf}
\caption{Abstract tree for ``Man uppskattar dem".}
\label{fig:mappMan}
\end{figure}
When translating a senctence like this, it is thus not possible to simply glue the
subject to the verb with \verb|PredVP| and the normal top-down? approach fails.\\
%Therefore, the translator must change the type of the 
%sentence when inside the SS.
